Darwin's Dream: Unifying ecological and evolutionary change

### 2025-05-31 response_to_reviewers.md

Overall, population change appears in models where the Price equation also appears, but what has always been missing is a derivation of both from the same starting point, from first principles of what actually happens in a population. Logically, this has to be possible because the same mechanisms that underlie population change (birth and death) also underlie evolutionary change. But up until now, to the best of our knowledge, the link between population and evolutionary change has always been carefully considered but ultimately ad hoc (as in the abovementioned references and, to our knowledge, all eco-evolutionary models). Given that we are looking at the same mechanisms of population change, we really should be able to demonstrate that both population change and evolutionary change can 'pop out' as intrinsically and necessarily linked from the same starting point.

Figuring this starting point out was not easy; it needed to be done deductively, which inevitably led to many dead ends. Initially, we reasoned that because the Price equation is sufficiently abstract to describe trait change in any population (including groups and higher moments of traits), its recursive properties, and how useful it is for making formal links across disciplines such as physics and information theory (Frank 2015; 2016), we should be able to use it directly to recover population density change. This ended up being a dead end for reasons we will not dwell on (briefly, conservation of total probability -- that entity frequencies need to sum to 1 -- precludes the Price equation from being used to directly to recover absolute population change -- it's not a simple matter of, e.g., setting 'z' to population density). We had different issues when starting from population change and trying to recover the Price equation. There are a lot of ways to make coherent models using these avenues of inquiry, but none that recover universal and exact descriptions of population and evolutionary change. The broader equation, our equation 1, is indeed necessary to derive exact and universal eco-evolutionary change from first principles; it thereby serves as a formal definition for eco-evolutionary change in any system. It perhaps appears simple and obvious in hindsight, as if it could be inferred quickly from existing models by anyone giving it a modicum of thought. But we believe this appearance is highly deceptive; if this formal link were trivial to discover, then someone would have surely pointed it out already. After much investigation, we are confident that no one has, and while there are fundamental equations defining ecological and evolutionary change, separately, our work here is unique in establishing a fundamental equation that defines eco-evolutionary change.

### 2025-05-31 foundations_of_ecoevo.Rmd
Fix the misspelling of Fisher's name

> **Box 2:** The Price equation is an abstract formula to represent evolutionary change. Formulated originally in the early 1970s by George Price [@Price1970; @Price1972], it postulates some basic properties that all evolutionary systems must satisfy: change over time, ancestor and descendant relations, and a character or phenotype [@Rice2004]. Using simple algebraic language, the Price equation represents evolutionary change with the predominant notation, $$\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).$$ In the above equation, $\Delta\bar{z}$ is the change in the average character value $z$ over a time step of arbitrary length, $w$ is an individual's absolute fitness, and $\bar{w}$ average population fitness. On the right-hand side of the equation, the first term is the covariance between a character value $z$ and fitness $w$, which reflects $\bar{z}$ change attributable to differential survival and reproduction. The second term is the expected value of $\Delta z$, which reflects the extent to which offspring deviate from parents in $z$ [@Rice2004; @Okasha2006; @Frank2012]. A more specific version of the covariance term was already known within the quantitative and population genetics tradition [@Robertson1966], usually representing the action of natural selection. The Price equation adds an expectation term and abstracts away from any specific mechanisms of replication or reproduction, or mechanisms of inheritance. Its definitional nature and lack of substantive biological assumptions has been portrayed both as a strength [@Rice2004; @Frank2012; @Luque2016; @Baravalle2022], and its greatest weakness. The abstract nature of the Price equation places it at the top of the hierarchy of fundamental theorems of evolution from which the rest (Robertson's theorem, Fischer's fundamental theorem, breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be easily derived [@Queller2017; @Lehtonen2016; @Lehtonen2018]. This abstractness is also key to developing a more general view of evolution [@Rice2020; @Luque2021; @Edelaar2023]. In contrast, some researchers consider the Price equation just a triviality (even tautological), and useless without further modelling assumptions [@VanVeelen2005; @VanVeelen2012]. The debate remains open [@VanVeelen2020; @Baravalle2024].

>>>>

> **Box 2:** The Price equation is an abstract formula to represent evolutionary change. Formulated originally in the early 1970s by George Price [@Price1970; @Price1972], it postulates some basic properties that all evolutionary systems must satisfy: change over time, ancestor and descendant relations, and a character or phenotype [@Rice2004]. Using simple algebraic language, the Price equation represents evolutionary change with the predominant notation, $$\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).$$ In the above equation, $\Delta\bar{z}$ is the change in the average character value $z$ over a time step of arbitrary length, $w$ is an individual's absolute fitness, and $\bar{w}$ average population fitness. On the right-hand side of the equation, the first term is the covariance between a character value $z$ and fitness $w$, which reflects $\bar{z}$ change attributable to differential survival and reproduction. The second term is the expected value of $\Delta z$, which reflects the extent to which offspring deviate from parents in $z$ [@Rice2004; @Okasha2006; @Frank2012]. A more specific version of the covariance term was already known within the quantitative and population genetics tradition [@Robertson1966], usually representing the action of natural selection. The Price equation adds an expectation term and abstracts away from any specific mechanisms of replication or reproduction, or mechanisms of inheritance. Its definitional nature and lack of substantive biological assumptions has been portrayed both as a strength [@Rice2004; @Frank2012; @Luque2016; @Baravalle2022], and its greatest weakness. The abstract nature of the Price equation places it at the top of the hierarchy of fundamental theorems of evolution from which the rest (Robertson's theorem, Fisher's fundamental theorem, breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be easily derived [@Queller2017; @Lehtonen2016; @Lehtonen2018]. This abstractness is also key to developing a more general view of evolution [@Rice2020; @Luque2021; @Edelaar2023]. In contrast, some researchers consider the Price equation just a triviality (even tautological), and useless without further modelling assumptions [@VanVeelen2005; @VanVeelen2012]. The debate remains open [@VanVeelen2020; @Baravalle2024].
